Hazelnut Live: Type-Theoretic Foundations for Running Incomplete Programs

### 2017-06-09 calculus.tex
fixed expand-ana-non-empty-hole
changing the rule names to match with the corresponding agda constructor names. this entails a change in naming convention, unfortunately: in agda, it's common to name functions with lower case and hyphens as in `subst-commute` or similar, and name constructors in the ML tradition with all caps as in `LCons`. so i've followed that convention.
roughly speaking, the first `n-1` letters in the all-caps prefix of a
constructor name denote the judgement it comes from, and the rest denotes
the form it applies to. if more than one rule applies to the same form,
then th

\begin{figure}[h]
\judgbox{\tconsistent{\htau_1}{\htau_2}}{$\htau_1$ is consistent with $\htau_2$}
\begin{mathpar}
\inferrule{ }{
  \tconsistent{\tehole}{\htau}
}

\inferrule{ }{
  \tconsistent{\htau}{\tehole}
}

\inferrule{ }{
  \tconsistent{\htau}{\htau}
}

\inferrule{
  \tconsistent{\htau_1}{\htau_1'}\\
  \tconsistent{\htau_2}{\htau_2'}
}{
  \tconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
}
%
% \inferrule{
%   \tconsistent{\htau_1}{\htau_1'}\\
%   \tconsistent{\htau_2}{\htau_2'}
% }{
%   \tconsistent{\tprod{\htau_1}{\htau_2}}{\tprod{\htau_1'}{\htau_2'}}
% }
%
% \inferrule{
%   \tconsistent{\htau_1}{\htau_1'}\\
%   \tconsistent{\htau_2}{\htau_2'}
% }{
%   \tconsistent{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1'}{\htau_2'}}
% }
\end{mathpar}

\judgbox{\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}}{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
\begin{mathpar}
\inferrule{ }{
  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
}

\inferrule{ }{
  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
}
\end{mathpar}

@@

% \inferrule{ }{
%   \summatch{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1}{\htau_2}}
% }
% \end{mathpar}
\caption{Type Consistency and Matched Types \todo{inconsistent}}
\label{fig:tconsistent}
\label{fig:arrmatch}
\end{figure}

\begin{figure}[h]
\judgbox{\hsyn{\hGamma}{\hexp}{\htau}}{$\htau$ synthesizes type $\htau$}
\begin{mathpar}
\inferrule[syn-const]{ }{
  \hsyn{\hGamma}{c}{b}
}

\inferrule[syn-var]{ }{
  \hsyn{\hGamma, x : \htau}{x}{\htau}
}

\inferrule[syn-alam]{
  \hsyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hsyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}

\inferrule[syn-ap]{
  \hsyn{\hGamma}{\hexp_1}{\htau_1}\\
  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
  \hana{\hGamma}{\hexp_2}{\htau_2}
}{
  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
}

\inferrule[syn-empty-hole]{ }{
  \hsyn{\hGamma}{\hehole{u}}{\tehole}
}

\inferrule[syn-non-empty-hole]{
  \hsyn{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}
}

\inferrule[syn-ascription]{
  \hana{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hexp : \htau}{\htau}
}
\end{mathpar}

\judgbox{\hana{\hGamma}{\hexp}{\htau}}{$\htau$ analyzes against type $\htau$}
\begin{mathpar}
\inferrule[ana-lam]{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
}

\inferrule[subsumption]{
  \hsyn{\hGamma}{\hexp}{\htau}\\
  \tconsistent{\htau}{\htau'}
}{
  \hana{\hGamma}{\hexp}{\htau'}
}
\end{mathpar}
\caption{Bidirectional Typing of H-Expressions}
\label{fig:tconsistent}
\label{fig:arrmatch}
\end{figure}

\begin{figure}[h]
\judgbox{\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}}{ }
\begin{mathpar}
\inferrule[expand-constant]{ }{
  \expandSyn{\hGamma}{c}{b}{c}{\emptyset}
}

\inferrule[expand-var]{ }{
  \expandSyn{\hGamma, x : \htau}{x}{\htau}{x}{\emptyset}
}

\inferrule[expand-alam]{
  \expandSyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\Delta}
}{
  \expandSyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\Delta}
}

\inferrule[expand-ap-1]{
  \hsyn{\hGamma}{\hexp_1}{\tehole}\\
  \expandAna{\hGamma}{\hexp_2}{\tehole}{\dexp_2}{\htau_2}{\Delta_2}\\
  \expandAna{\hGamma}{\hexp_1}{\tarr{\htau_2}{\tehole}}{\dexp_1}{\htau_1}{\Delta_1}
}{
  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}{\hap{(\dcast{\dexp_1}{\tarr{\htau_2}{\tehole}})}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
}

\inferrule[expand-ap-2]{
  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau'_2}{\Delta_2}\\
  \htau_2 \neq \htau'_2
}{
  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dcast{\dexp_2}{\htau_2}}}{\Dunion{\Delta_1}{\Delta_2}}
}

\inferrule[expand-ap-3]{
  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau_2}{\Delta_2}
}{
  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
}\\
%
% \inferrule[expand-pair]{
%   \expandSyn{\hGamma}{\hexp_1}{\htau_1}{\dexp_1}{\Delta_1}\\
%   \expandSyn{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\Delta_2}
% }{
%   \expandSyn{\hGamma}{\hpair{\hexp_1}{\hexp_2}}{\tprod{\htau_1}{\htau_2}}{\hpair{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
% }
%
% \inferrule[expand-prj]{
%   a
% }{
%   b
% }
%
% (inj)

%
% \inferrule[expand-plus]{ }{
%   \expandSyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}{\hadd{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
% }
%
\inferrule[expand-syn-empty-hole]{ }{
  \expandSyn{\hGamma}{\hehole{u}}{\tehole}{\dehole{u}{\idof{\hGamma}}}{\Dbinding{u}{\hGamma}{\tehole}}
}

\inferrule[expand-syn-non-empty-hole]{
  \expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}
}{
  \expandSyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\Delta, \Dbinding{u}{\hGamma}{\tehole}}
}\\
%
\inferrule[expand-asc-1]{
  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}\\
  \htau \neq \htau'
}{
  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dcast{\dexp}{\htau}}{\Delta}
}

\inferrule[expand-asc-2]{
  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}
}{
  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dexp}{\Delta}
}
\end{mathpar}

\judgbox{\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}}{ }
\begin{mathpar}
\inferrule[expand-lam]{
  \expandAna{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\htau'_2}{\Delta}
}{
  \expandAna{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2'}}{\Delta}
}

\inferrule[expand-subsume]{
  \hexp \neq \hehole{u}\\
  \hexp \neq \hhole{\hexp'}{u}\\\\
  \expandSyn{\hGamma}{\hexp}{\htau'}{\dexp}{\Delta}\\
  \tconsistent{\htau}{\htau'}
}{
  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}
}

\inferrule[expand-ana-empty-hole]{ }{
  \expandAna{\hGamma}{\hehole{u}}{\htau}{\dehole{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
}

\inferrule[expand-ana-non-empty-hole]{ }{
  \expandAna{\hGamma}{\hhole{\hexp}{u}}{\htau}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
}
\end{mathpar}
\caption{Algorithmic Expansion}
\label{fig:expandSyn}
\label{fig:expandAna}
\end{figure}

\begin{figure}[h!]
\judgbox{\hasType{\Delta}{\hGamma}{\dexp}{\htau}}{$\dexp$ is assigned type $\htau$}
\begin{mathpar}
\inferrule[ty-assign-const]{ }{
  \hasType{\Delta}{\hGamma}{c}{b}
}

\inferrule[ty-assign-var]{ }{
	\hasType{\Delta}{\hGamma, x : \htau}{x}{\htau}
}

\inferrule[ty-assign-lam]{
  \hasType{\Delta}{\hGamma, x : \htau_1}{\dexp}{\htau_2}
}{
  \hasType{\Delta}{\hGamma}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2}}
}

\inferrule[ty-assign-ap]{
  \hasType{\Delta}{\hGamma}{\dexp_1}{\htau_1}\\
  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
  \hasType{\Delta}{\hGamma}{\dexp_2}{\htau_2}
}{
  \hasType{\Delta}{\hGamma}{\hap{\dexp_1}{\dexp_2}}{\htau}
}

\inferrule[ty-assign-empty-hole]{
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dehole{u}{\sigma}}{\htau}
}

\inferrule[ty-assign-non-empty-hole]{
  \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dhole{\dexp}{u}{\sigma}}{\htau}
}

\inferrule[ty-assign-cast]{
  \hasType{\Delta}{\Gamma}{\dexp}{\htau'}\\
  \tconsistent{\htau}{\htau'}
}{
  \hasType{\Delta}{\hGamma}{\dcast{\dexp}{\htau}}{\htau}
}
\end{mathpar}
\caption{Type Assignment for Dynamic H-Expressions}
\label{fig:hasType}
\end{figure}

\begin{definition}[Substitution Type Assignment]
$\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
\end{definition}

\begin{figure}[h!]
\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
\[
\begin{array}{lcl}
[\dexp_1 / x] c
&=&
c\\
{[\dexp_1 / x] x}
&=&
\dexp_1\\
{[\dexp_1 / x] y}
&=&
y\\
{[\dexp_1 / x] \dlam{y}{\dexp_2}}
&=&
{\dlam{y}{[\dexp_1 / x] \dexp_2}}
\\
{[\dexp_1 / x] \dap{e_2}{e_3}}
&=&
{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
\\
% {[\dexp_1 / x] \dinj{i}{e_2}}
% &=&
% {\dinj{i}{[\dexp_1 / x] e_2}}
% \\
% {[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
% &=&
% {\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
% \\
{[\dexp_1 / x]} \dehole{\mvar}{\subst}
&=&
\dehole{\mvar}{[\dexp_1 / x] \subst}
\\
{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}}
&=& 
\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}
\end{array}
\]
\end{figure}

\begin{figure}
\judgbox{\isValue{\dexp}}{$\dexp$ is a value}
\begin{mathpar}
\inferrule[v-constant]{ }{
  \isValue{c}
}

\inferrule[v-lam]{ }{
  \isValue{\halam{x}{\htau}{\dexp}}
}
\end{mathpar}

\judgbox{\isIndet{\dexp}}{$\dexp$ is indeterminate}
\begin{mathpar}
\inferrule[indet-empty-hole]
{ }
{\isIndet{\dehole{\mvar}{\subst}}}

\inferrule[indet-non-empty-hole]
{\isFinal{\dexp}}
{\isIndet{\dhole{\dexp}{\mvar}{\subst}}}

\inferrule[indet-app]
{\isIndet{\dexp_1}\\
% \isFinal{\dexp_2}~\text{\todo{??}}}
 \isFinal{\dexp_2}}
{\isIndet{\dap{\dexp_1}{\dexp_2}}}

\inferrule[indet-cast]
{\isIndet{\dexp}}
{\isIndet{\dcast{\dexp}{\htau}}}

@@

\judgbox{\isErr{\Delta}{\dexp}}{Under $\Delta$, $\dexp$ contains a cast error}
\begin{mathpar}
\inferrule[err-nemp-hole~\todo{??}]
{\subst \sim_? \Gamma_2
\\ 
\isErr{\Delta; \Gamma_1, \Gamma_2}{\dexp}}
{\isErr{\Delta; \Gamma_1}{\dhole{\dexp}{\mvar}{\subst}}}

\inferrule[err-cast-error~\todo{??}]
% XXX use macro type assignment judgement here
{\hasType{\Delta}{\todo{\emptyset}}{e}{\htau_2}\\
 \tinconsistent{\htau_1}{\htau_2}}
{\isErr{\Delta}{\dcast{\dexp}{\htau_1}}}

\inferrule[err-lam~\todo{??}]
% XXX use macro type assignment judgement here
{\isErr{\Delta; \Gamma, x:\htau_1}{\dexp}}
{\isErr{\Delta; \Gamma}{\halam{x}{\htau}{\dexp}}}

@@

\inferrule[err-app-1]
{
 \isErr{\Delta}{\dexp_1}
 %\\\
 %isFinal{\dexp_2}~\text{\todo{??}}
}
{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}

>>>>

\begin{figure}[h]
\judgbox{\tconsistent{\htau_1}{\htau_2}}{$\htau_1$ is consistent with $\htau_2$}
\begin{mathpar}
\inferrule[TCHole1]{ }{
  \tconsistent{\tehole}{\htau}
}

\inferrule[TCHole2]{ }{
  \tconsistent{\htau}{\tehole}
}

\inferrule[TCRefl]{ }{
  \tconsistent{\htau}{\htau}
}

\inferrule[TCArr]{
  \tconsistent{\htau_1}{\htau_1'}\\
  \tconsistent{\htau_2}{\htau_2'}
}{
  \tconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1'}{\htau_2'}}
}
%
% \inferrule{
%   \tconsistent{\htau_1}{\htau_1'}\\
%   \tconsistent{\htau_2}{\htau_2'}
% }{
%   \tconsistent{\tprod{\htau_1}{\htau_2}}{\tprod{\htau_1'}{\htau_2'}}
% }
%
% \inferrule{
%   \tconsistent{\htau_1}{\htau_1'}\\
%   \tconsistent{\htau_2}{\htau_2'}
% }{
%   \tconsistent{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1'}{\htau_2'}}
% }
\end{mathpar}

\judgbox{\tinconsistent{\htau}{\htau'}}{$\htau$ is inconsistent with $\htau'$}
\begin{mathpar}
    \inferrule[ICBaseArr1]{ }{
      \tinconsistent{\tb}{\tarr{\htau_1}{\htau_2}}
    }

    \inferrule[ICBaseArr2]{ }{
      \tinconsistent{\tarr{\htau_1}{\htau_2}}{\tb}
    }

    \inferrule[ICArr1]{
      \tinconsistent{\htau_1}{\htau_3}
    }{
      \tinconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_3}{\htau_4}}
    }

    \inferrule[ICArr2]{
      \tinconsistent{\htau_2}{\htau_4}
    }{
      \tinconsistent{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_3}{\htau_4}}
    }
\end{mathpar}

\judgbox{\arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}}{$\htau$ has matched arrow type $\tarr{\htau_1}{\htau_2}$}
\begin{mathpar}
\inferrule[MAHole]{ }{
  \arrmatch{\tehole}{\tarr{\tehole}{\tehole}}
}

\inferrule[MAArr]{ }{
  \arrmatch{\tarr{\htau_1}{\htau_2}}{\tarr{\htau_1}{\htau_2}}
}
\end{mathpar}

@@

% \inferrule{ }{
%   \summatch{\tsum{\htau_1}{\htau_2}}{\tsum{\htau_1}{\htau_2}}
% }
% \end{mathpar}
\caption{Type Consistency, Inconsistency, and Matching}
\label{fig:tconsistent}
\label{fig:arrmatch}
\end{figure}

\begin{figure}[h]
\judgbox{\hsyn{\hGamma}{\hexp}{\htau}}{$\htau$ synthesizes type $\htau$}
\begin{mathpar}
\inferrule[SConst]{ }{
  \hsyn{\hGamma}{c}{b}
}

\inferrule[SVar]{ }{
  \hsyn{\hGamma, x : \htau}{x}{\htau}
}

\inferrule[SLam]{
  \hsyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hsyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}
}

\inferrule[SAp]{
  \hsyn{\hGamma}{\hexp_1}{\htau_1}\\
  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
  \hana{\hGamma}{\hexp_2}{\htau_2}
}{
  \hsyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}
}

\inferrule[SEHole]{ }{
  \hsyn{\hGamma}{\hehole{u}}{\tehole}
}

\inferrule[SNEHole]{
  \hsyn{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}
}

\inferrule[SAsc]{
  \hana{\hGamma}{\hexp}{\htau}
}{
  \hsyn{\hGamma}{\hexp : \htau}{\htau}
}
\end{mathpar}

\judgbox{\hana{\hGamma}{\hexp}{\htau}}{$\htau$ analyzes against type $\htau$}
\begin{mathpar}
\inferrule[ALam]{
  \arrmatch{\htau}{\tarr{\htau_1}{\htau_2}}\\
  \hana{\hGamma, x : \htau_1}{\hexp}{\htau_2}
}{
  \hana{\hGamma}{\hlam{x}{\hexp}}{\htau}
}

\inferrule[ASubsume]{
  \hsyn{\hGamma}{\hexp}{\htau}\\
  \tconsistent{\htau}{\htau'}
}{
  \hana{\hGamma}{\hexp}{\htau'}
}
\end{mathpar}
\caption{Bidirectional Typing of H-Expressions}
\label{fig:tconsistent}
\label{fig:arrmatch}
\end{figure}

\todo{definition of complete types and terms? do we care about completeness
  of $\hexp$s or $\dexp$s?}

\begin{figure}[h]
\judgbox{\expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}}{ }
\begin{mathpar}
\inferrule[ESConst]{ }{
  \expandSyn{\hGamma}{c}{b}{c}{\emptyset}
}

\inferrule[ESVar]{ }{
  \expandSyn{\hGamma, x : \htau}{x}{\htau}{x}{\emptyset}
}

\inferrule[ESLam]{
  \expandSyn{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\Delta}
}{
  \expandSyn{\hGamma}{\halam{x}{\htau_1}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\Delta}
}

\inferrule[ESAp1]{
  \hsyn{\hGamma}{\hexp_1}{\tehole}\\
  \expandAna{\hGamma}{\hexp_2}{\tehole}{\dexp_2}{\htau_2}{\Delta_2}\\
  \expandAna{\hGamma}{\hexp_1}{\tarr{\htau_2}{\tehole}}{\dexp_1}{\htau_1}{\Delta_1}
}{
  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\tehole}{\hap{(\dcast{\dexp_1}{\tarr{\htau_2}{\tehole}})}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
}

\inferrule[ESAp2]{
  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau'_2}{\Delta_2}\\
  \htau_2 \neq \htau'_2
}{
  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dcast{\dexp_2}{\htau_2}}}{\Dunion{\Delta_1}{\Delta_2}}
}

\inferrule[ESAp3]{
  \expandSyn{\hGamma}{\hexp_1}{\tarr{\htau_2}{\htau}}{\dexp_1}{\Delta_1}\\
  \expandAna{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\htau_2}{\Delta_2}
}{
  \expandSyn{\hGamma}{\hap{\hexp_1}{\hexp_2}}{\htau}{\hap{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
}\\
%
% \inferrule[expand-pair]{
%   \expandSyn{\hGamma}{\hexp_1}{\htau_1}{\dexp_1}{\Delta_1}\\
%   \expandSyn{\hGamma}{\hexp_2}{\htau_2}{\dexp_2}{\Delta_2}
% }{
%   \expandSyn{\hGamma}{\hpair{\hexp_1}{\hexp_2}}{\tprod{\htau_1}{\htau_2}}{\hpair{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
% }
%
% \inferrule[expand-prj]{
%   a
% }{
%   b
% }
%
% (inj)

%
% \inferrule[expand-plus]{ }{
%   \expandSyn{\hGamma}{\hadd{\hexp_1}{\hexp_2}}{\tnum}{\hadd{\dexp_1}{\dexp_2}}{\Dunion{\Delta_1}{\Delta_2}}
% }
%
\inferrule[ESEHole]{ }{
  \expandSyn{\hGamma}{\hehole{u}}{\tehole}{\dehole{u}{\idof{\hGamma}}}{\Dbinding{u}{\hGamma}{\tehole}}
}

\inferrule[ESNEHole]{
  \expandSyn{\hGamma}{\hexp}{\htau}{\dexp}{\Delta}
}{
  \expandSyn{\hGamma}{\hhole{\hexp}{u}}{\tehole}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\Delta, \Dbinding{u}{\hGamma}{\tehole}}
}\\
%
\inferrule[ESAsc1]{
  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}\\
  \htau \neq \htau'
}{
  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dcast{\dexp}{\htau}}{\Delta}
}

\inferrule[ESAsc2]{
  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}
}{
  \expandSyn{\hGamma}{\hexp : \htau}{\htau}{\dexp}{\Delta}
}
\end{mathpar}

\judgbox{\expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau}{\Delta}}{ }
\begin{mathpar}
\inferrule[EALam]{
  \expandAna{\hGamma, x : \htau_1}{\hexp}{\htau_2}{\dexp}{\htau'_2}{\Delta}
}{
  \expandAna{\hGamma}{\hlam{x}{\hexp}}{\tarr{\htau_1}{\htau_2}}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2'}}{\Delta}
}

\inferrule[EASubsume]{
  \hexp \neq \hehole{u}\\
  \hexp \neq \hhole{\hexp'}{u}\\\\
  \expandSyn{\hGamma}{\hexp}{\htau'}{\dexp}{\Delta}\\
  \tconsistent{\htau}{\htau'}
}{
  \expandAna{\hGamma}{\hexp}{\htau}{\dexp}{\htau'}{\Delta}
}

\inferrule[EAEHole]{ }{
  \expandAna{\hGamma}{\hehole{u}}{\htau}{\dehole{u}{\idof{\hGamma}}}{\htau}{\Dbinding{u}{\hGamma}{\htau}}
}

\inferrule[EANEHole]{
  \expandSyn{\hGamma}{\hexp}{\htau'}{\dexp}{\Delta}\\
  \tconsistent{\htau}{\htau'}
}{
  \expandAna{\hGamma}{\hhole{\hexp}{u}}{\htau}{\dhole{\dexp}{u}{\idof{\hGamma}}}{\htau}{\Delta, \Dbinding{u}{\hGamma}{\htau}}
}
\end{mathpar}
\caption{Algorithmic Expansion}
\label{fig:expandSyn}
\label{fig:expandAna}
\end{figure}

\begin{figure}[!ht]
\judgbox{\hasType{\Delta}{\hGamma}{\dexp}{\htau}}{$\dexp$ is assigned type $\htau$}
\begin{mathpar}
\inferrule[TAConst]{ }{
  \hasType{\Delta}{\hGamma}{c}{b}
}

\inferrule[TAVar]{ }{
	\hasType{\Delta}{\hGamma, x : \htau}{x}{\htau}
}

\inferrule[TALam]{
  \hasType{\Delta}{\hGamma, x : \htau_1}{\dexp}{\htau_2}
}{
  \hasType{\Delta}{\hGamma}{\halam{x}{\htau_1}{\dexp}}{\tarr{\htau_1}{\htau_2}}
}

\inferrule[TAAp]{
  \hasType{\Delta}{\hGamma}{\dexp_1}{\htau_1}\\
  \arrmatch{\htau_1}{\tarr{\htau_2}{\htau}}\\\\
  \hasType{\Delta}{\hGamma}{\dexp_2}{\htau_2}
}{
  \hasType{\Delta}{\hGamma}{\hap{\dexp_1}{\dexp_2}}{\htau}
}

\inferrule[TAEHole]{
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dehole{u}{\sigma}}{\htau}
}

\inferrule[TANEHole]{
  \hasType{\Delta}{\hGamma}{\dexp}{\htau'}\\
  \hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}
}{
  \hasType{\Delta, \Dbinding{u}{\hGamma'}{\htau}}{\hGamma}{\dhole{\dexp}{u}{\sigma}}{\htau}
}

\inferrule[TACast]{
  \hasType{\Delta}{\Gamma}{\dexp}{\htau'}\\
  \tconsistent{\htau}{\htau'}
}{
  \hasType{\Delta}{\hGamma}{\dcast{\dexp}{\htau}}{\htau}
}
\end{mathpar}
\caption{Type Assignment for Dynamic H-Expressions}
\label{fig:hasType}
\end{figure}

\begin{figure}[!ht]
  \begin{definition}
    $\hasType{\Delta}{\hGamma}{\sigma}{\hGamma'}$ iff for each $\dexp/x \in \sigma$, we have $x : \htau \in \hGamma'$ and $\hasType{\Delta}{\hGamma}{\dexp}{\htau}$.
  \end{definition}
  \caption{substitution type assignment}
\end{figure}

\begin{figure}[!ht]
\judgbox{[\dexp_1 / x] \dexp_2 = \dexp_3}{Substitution}
\[
\begin{array}{lcl}
[\dexp_1 / x] c
&=&
c\\
{[\dexp_1 / x] x}
&=&
\dexp_1\\
{[\dexp_1 / x] y}
&=&
y\\
{[\dexp_1 / x] \dlam{y}{\htau}{\dexp_2}}
&=&
{\dlam{y}{\htau}{[\dexp_1 / x] \dexp_2}}
\\
{[\dexp_1 / x] \dap{e_2}{e_3}}
&=&
{\dapP{[\dexp_1 / x] e_2}{[\dexp_1 / x] e_3}}
\\
% {[\dexp_1 / x] \dinj{i}{e_2}}
% &=&
% {\dinj{i}{[\dexp_1 / x] e_2}}
% \\
% {[\dexp_1 / x] \dcase{e}{x}{e_x}{y}{e_y}}
% &=&
% {\dcase{[\dexp_1 / x]e}{x}{[\dexp_1 / x]e_x}{y}{[\dexp_1 / x]e_y}}
% \\
{[\dexp_1 / x]} \dehole{\mvar}{\subst}
&=&
\dehole{\mvar}{[\dexp_1 / x] \subst}
\\
{[\dexp_1 / x] \dhole{\dexp_2}{\mvar}{\subst}}
&=&
\dhole{ [ \dexp_1 / x ] \dexp_2 }{\mvar}{[\dexp_1 / x] \subst}
\end{array}
\]
\end{figure}

\begin{figure}[!ht]
\judgbox{\isValue{\dexp}}{$\dexp$ is a value}
\begin{mathpar}
\inferrule[VConst]{ }{
  \isValue{c}
}

\inferrule[VLam]{ }{
  \isValue{\halam{x}{\htau}{\dexp}}
}
\end{mathpar}

\judgbox{\isIndet{\dexp}}{$\dexp$ is indeterminate}
\begin{mathpar}
\inferrule[IEHole]
{ }
{\isIndet{\dehole{\mvar}{\subst}}}

\inferrule[INEHole]
{\isFinal{\dexp}}
{\isIndet{\dhole{\dexp}{\mvar}{\subst}}}

\inferrule[IAp]
{\isIndet{\dexp_1}\\
% \isFinal{\dexp_2}~\text{\todo{??}}}
 \isFinal{\dexp_2}}
{\isIndet{\dap{\dexp_1}{\dexp_2}}}

\inferrule[ICast]
{\isIndet{\dexp}}
{\isIndet{\dcast{\dexp}{\htau}}}

@@

\judgbox{\isFinal{\dexp}}{$\dexp$ is final (it cannot be stepped)}
\begin{mathpar}
\Infer{FVal}
{\isValue{\dexp}}{\isFinal{\dexp}}
\and
\Infer{FIndet}
{\isIndet{\dexp}}{\isFinal{\dexp}}
%% No longer a rule: Errors are not indet anymore:
%% \and
%% \Infer{f-err}
%% {\isErr{\dexp}}{\isFinal{\dexp}}
\end{mathpar}

\judgbox{\isErr{\Delta}{\dexp}}{Under $\Delta$, $\dexp$ contains a cast error}
\begin{mathpar}
\inferrule[ERNEHole~\todo{??}]
{\subst \sim_? \Gamma_2
\\
\isErr{\Delta; \Gamma_1, \Gamma_2}{\dexp}}
{\isErr{\Delta; \Gamma_1}{\dhole{\dexp}{\mvar}{\subst}}}

\inferrule[ERCastError~\todo{??}]
% XXX use macro type assignment judgement here
{\hasType{\Delta}{\todo{\emptyset}}{e}{\htau_2}\\
 \tinconsistent{\htau_1}{\htau_2}}
{\isErr{\Delta}{\dcast{\dexp}{\htau_1}}}

\inferrule[ERLam~\todo{??}]
% XXX use macro type assignment judgement here
{\isErr{\Delta; \Gamma, x:\htau_1}{\dexp}}
{\isErr{\Delta; \Gamma}{\halam{x}{\htau}{\dexp}}}

@@

\inferrule[ERAp1]
{
 \isErr{\Delta}{\dexp_1}
 %\\\
 %isFinal{\dexp_2}~\text{\todo{??}}
}
{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}

\inferrule[ERAp2]
{
 %\isIndet{\dexp_1}
 %\\
 \isErr{\Delta}{\dexp_2}
}
{\isErr{\Delta}{\dap{\dexp_1}{\dexp_2}}}
